---
title: "Classwork 4: Data manipulation and plotting"
format:
  html:
    toc: true
    df-print: kable
    code-tools: true
    embed-resources: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```


```{r}
library(tidyverse)
library(tidycensus)
url<-'https://dataverse.harvard.edu/api/access/datafile/12027896'
ccc<-read_csv(url)

```



# Question 1


Create a regular expression to extract events related to Israel/Palestine issues. Calculate the number of events on this issue per week and create a line plot to show the count of events.

(make sure your line plot includes "0"s for weeks that have no data. You could either use a cross join or the `complete` function to accomplish this)

```{r}
# Code here


```


# Question 2

Use a series of joins to get a single dataset that allows us to see the association between county population and the number of protest events since January 2025. Make a scatter plot to visualize the relationship between population and protest frequency.




We can use data from the US census bureau, which we can access using the tidycensus package:

```{r}

census_data <- get_acs(
  geography = "county",
  variables = "B01001_001", # population
    year = 2023)



```


```{r}
# code here

```

# Question 3

Get a data set that contains a list of all counties that haven't had any protests during the time period covered.


```{r}
# Code here

```


# Question 4

The `organizations` column contains one or more organizations involved in each event. Make a list of the top 10 organizations by number of protests and then use ggplot and `graph_bar` to visualize the results.


```{r}
# Code here

```

